---
title: "Ec/CS/ACM 112 - PS7"
author: "Ben Juarez"
output:
  pdf_document: default
  html_document: default
---

# Question 2

## prelims
```{r}
rm(list=ls())
set.seed(123)
library(Rcpp)
library(rstan)
library(bayesplot)
options(mc.cores = parallel::detectCores())
rstan_options(auto_write = TRUE)

data = read.csv("~/Desktop/MetabolicRate.csv")
Mrate = data$Mrate
BodySize = data$BodySize
logMrate = log(Mrate)
logBodySize = log(BodySize)
Instar = data$Instar
N = length(data[,1])
```

## Step 2
```{r}
num_iter = 10000
fit = stan("Ec122_PS7.stan", iter = num_iter, chains = 4, 
           data = list(log_Mrate=logMrate,
                       log_BodySize=logBodySize,
                       Instar=Instar,
                       N=N))
print(fit,probs = c(0.25,0.5,0.75))
```

## step 3
```{r}
traceplot(fit, pars = c("beta0", "beta1", "beta2", "sigma"), 
          inc_warmup = FALSE, nrow = 4, window = c(num_iter-2000, num_iter))
```


## step 4
```{r}
pairs(fit, pars = c("beta0", "beta1", "beta2", "sigma", "lp__"))
```

## step 6
```{r}
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
y = logMrate
ppc_dens_overlay(logMrate, yrep[indices,])
```

It would appear that this simple linear regression model does indeed provide a satisfactory account of the data considering the results of this posterior predictive check.  Looking at the y and y_rep curves, we see that the y_rep curves follow the general shape of y quite well.  However, we perhaps see some slight inaccuracies with higher Mrate levels around 0.  Although, especially considering that we are given 305 observations, we see that the model fits quite well. 


## step 7
```{r}
library(gridExtra)
new_data = subset(data, Instar==1)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p1 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

new_data = subset(data, Instar==2)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p1 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

new_data = subset(data, Instar==2)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p2 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

new_data = subset(data, Instar==3)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p3 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

new_data = subset(data, Instar==4)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p4 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

new_data = subset(data, Instar==5)
fit = stan("Ec122_PS7.stan", iter = 10000, chains = 4,
         data = list(log_Mrate=log(new_data$Mrate),
                     log_BodySize=log(new_data$BodySize),
                     Instar=new_data$Instar,
                     N=length(new_data[,1])))
yrep = as.matrix(fit, pars = "yPred")
indices = sample(nrow(yrep),100)
p5 = ppc_dens_overlay(log(new_data$Mrate), yrep[indices,])

grid.arrange(p1, p2, p3, p4, p5, nrow = 3)
```

Overall, it does not appear that there is much of a difference in the ability of the model to predict the data at different stages of the worms' life cycle.  If we look more closely, one might say that Instar=1 and Instar=5 have a slightly weakened impact on the model.  For Instar=1, the y_rep curves tend to be a bit higher than they should around -0.5 and 0.  For Instar=5, the y_rep curves seem to be a little too generalized around the 3 to 3.5 range considering that y rises and falls in this range while the y_rep curves seem to just be continually increasing instead.
